---
title: "NASA_API"
author: "Isabella Westlauer"
date: "2026-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#NASA API Exploratory Data Analysis

## Import libraries
```{r}
library(httr2)
# r library for making http requests to an API
library(tidyverse) 
# tidyverse includes tibble, dplyr, ggplot2, tidyr, readr, stringr, etc.
```

## API GET Request from SBDB
### Make an API GET request to NASA's Small-Body Database query API
#### We request three fields:
####  - spkid: unique identifier for the object
####  - rms: root-mean-square residual of the orbit fit
####  - last_obs: date of the most recent observation

```{r}
# Make an API GET request to sbdb_query.api

fields <- "rms,data_arc,spkid,last_obs,n_obs_used"
limit <- 1000 # limit the number of records returned

resp <- request("https://ssd-api.jpl.nasa.gov/sbdb_query.api") |>
  req_url_query(
    fields = fields,
    limit = limit) |>
  # Prevent the request from stopping execution on HTTP errors
  req_error(is_error = function(resp) FALSE) |>
  req_perform()
# Check HTTP status code:
# 200 indicates a successful request
resp$status_code
```
# Tidy the data
```{r}
# Parse the JSON response into an R list
data <- resp_body_json(resp)
# Inspect the field names returned by the API
str(data$fields, max.level = 2)

# The API returns the data as a list of rows (one list per object).
# do.call(rbind, ...) binds all rows together into a single matrix-like object.
# This is necessary to convert the nested JSON structure into rectangular data.
df <- as_tibble(
  do.call(rbind, data$data),
  .name_repair = "minimal"
)

# Assign column names using the metadata returned by the API
colnames(df) <- data$fields

# Some columns are stored as lists due to missing values in the API response.
# unlist() flattens each column into a standard atomic vector,
# which is required before converting data types.

df <- df |>
  mutate(
    spkid = unlist(spkid),
    rms = unlist(rms),
    last_obs = unlist(last_obs),
    data_arc = unlist(data_arc),
    n_obs_used = unlist(n_obs_used)
  )

# Preview a few values of last_obs to inspect formatting
# unique(df$last_obs)[1:10]
# unique(df$n_obs_used)[1:10]
unique(df$data_arc)[1:10]
```

## More data tidying
```{r}
# Convert columns to appropriate data types:
#  - spkid remains character
#  - rms is numeric
#  - n_obs_used is numeric
#  - data_arc is numeric
#  - last_obs is converted to Date
# Missing Data: 
# Empty strings in last_obs are replaced with NA before conversion
# compare to an ifna() in Excel
df <- df |>
  mutate(
    spkid = as.character(spkid),
    rms = as.numeric(rms),
    n_obs_used = as.numeric(n_obs_used),
    data_arc = as.numeric(data_arc),
    last_obs = na_if(last_obs, ""),
    last_obs = as.Date(last_obs, format = "%Y-%m-%d")
  )

# Inspect the cleaned data
head(df)

# Export the cleaned data to a CSV file
# write.csv(df, file = "output.csv", row.names = FALSE)
```

## Plot Distribution of Orbit RMS
### The RMS of the orbit of a small body measures how well the computed orbit fits the observational data. Lower RMS values indicate a better orbit fit.

```{r}
ggplot(df, aes(x = rms)) +
  geom_histogram(binwidth = 0.02, fill = "blue", color = "black") +
  labs(title = "Distribution of Orbit RMS values of Small Bodies",
       x = "RMS",
       y = "Count") +
  theme_minimal()
```
##What are the outliers? Are there objects with a relatively low RMS because of when they were last observed, or how recently they were observed?

##Plot last_obs vs count of objects
```{r}

plot(df$last_obs)
rms_cutoff <- quantile(df$rms, 0.05, na.rm = TRUE)
low_rms <- df %>%
  filter(rms <= rms_cutoff)
summary(low_rms$last_obs)
```
##A couple of these objects were last observed in 2021, the majority have been observed between 2024-2026.

##Compare the number of observations to the rms
```{r}
ggplot(df, aes(x = data_arc, y = rms)) +
  geom_point(alpha = 0.5) +
  scale_y_log10() +
  labs(
    title = "RMS vs Observation Arc Length",
    x = "Observation Arc Length (days)",
    y = "RMS (log scale)"
  )
```
```{r}
ggplot(df, aes(x = last_obs, y = rms)) +
  geom_point(alpha = 0.5) +
  scale_y_log10() +
  labs(
    title = "RMS vs Last Observation Date",
    x = "Last Observation Date",
    y = "RMS (log scale)"
  )
```


#### Observation arc length reflects the total time span between the first and most recent observations, not the age of the most recent measurement. Objects with arcs exceeding 50,000 days were typically discovered in the 19th century and have been re-observed repeatedly with modern instrumentation, resulting in highly constrained orbital solutions and very low rms values.

```{r}
rms_cutoff <- quantile(df$rms, 0.05, na.rm = TRUE)

ggplot(df, aes(x = data_arc, y = rms)) +
  geom_point(alpha = 0.4) +
  geom_point(
    data = subset(df, rms <= rms_cutoff),
    color = "red",
    alpha = 0.7
  ) +
  scale_y_log10() +
  labs(
    title = "RMS vs Observation Arc Length (Low RMS Highlighted)",
    x = "Observation Arc Length (days)",
    y = "RMS (log scale)"
  )
```




