---
title: "Web_Scraping_XML"
author: "Isabella Westlauer"
date: "2026-01-24"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(xml2)
library(dplyr)
library(ggplot2)
```

## Ingest the XML Document
```{r}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Frestaurants.xml"
doc <- read_xml(url)
```

## Extract the Nodes from the XML
```{r}
# view page source to find the names of the nodes
names <- xml_text(xml_find_all(doc, "//name"))
#so large it doesn't output to the console
zipcodes <- xml_text(xml_find_all(doc, "//zipcode")) 
#too large to output to console
council_districts <- xml_text(xml_find_all(doc, "//councildistrict")) 
```
##Create a data frame
```{r}
restaurants = data.frame(names, zipcodes, council_districts)
head(restaurants)
```

##Filter data to names including the word "sushi"
```{r}
# Are there any sushi restaurants in Baltimore? (all the data is in Baltimore)
# NOTE the string to search is case-sensitive by default
sushi_df <- restaurants %>%
  filter(grepl("SUSHI", names), ignore.case = TRUE)
length(sushi_df$names)
sushi_df <- restaurants %>%
  filter(grepl("sushi", names), ignore.case = TRUE)
length(sushi_df$names)

```
##Find downtown sushi restaurants, nested filters  
To identify sushi restaurants downtown, the dataset was filtered by council district and restaurant name. Council District 11 corresponds to the downtown area. The grepl() function was used with ignore.case = TRUE to identify restaurants whose names contain the word “sushi,” regardless of capitalization.
```{r}
downtown_restaurants <- filter(restaurants, council_districts == "11")
sushi_downtown <- downtown_restaurants %>%
  filter(grepl("sushi", names, ignore.case = TRUE))
length(sushi_downtown$names)
```
##A better way to do it, combined conditions in one filter
```{r}

sushi_downtown <- restaurants %>%
  filter(
    council_districts == "11",
    grepl("sushi", names, ignore.case = TRUE)
  )
# length(sushi_downtown$names)
# Name(s) of downtown sushi restaurant(s)
sushi_downtown$names
```

##Number of restaurants in each council district
```{r}
#count of names per council_district
district_counts <- restaurants %>%
  group_by(council_districts) %>%
  summarise(restaurant_count = n())
```


Council district values were originally stored as character data, which caused alphabetical ordering on the x-axis. Converting the variable to numeric (or specifying factor levels) ensures districts are displayed in numerical order.

```{r}

district_counts <- district_counts %>%
  mutate(council_districts = as.numeric(council_districts))

ggplot(district_counts, aes(
  x = council_districts, 
  y = restaurant_count, 
  label = restaurant_count)) +
  geom_col(fill = "blue") +
  geom_label(aes(label = restaurant_count),
             nudge_x = 0.25, nudge_y = 0.25,
             fill = "lightblue", color = "black") +


  scale_x_continuous(breaks = 1:14) +
  labs(
    title = "Number of Restaurants by Council District",
    x = "Council District",
    y = "Number of Restaurants"
  ) +
  theme_minimal()
```

Now I wonder what district 1 is? district 11 is downtown so it will have a lot of restaurants, but why does district 1 have so many?