---
title: "case study 1_2"
author: "Isabella Westlauer"
date: "2026-03-07"
output: html_document
---
---
title: "Case Study 1 - Attrition"
author: "Isabella Westlauer"
date: "2026-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
### libraries for EDA and Initial Modeling
library(skimr)
library(gt)
library(tidyverse)
library(DataExplorer)
library(ggplot2)
library(plotly)
library(GGally)
```
## Data intake
```{r}
# raw_data <- read.csv(file.choose())
raw_data <- read.csv("CaseStudy1-data.csv")
```

## Prior Probability of Attrition
```{r}
# Count and Percent of Attrition from original dataset
overall_rate <- mean(raw_data$Attrition == "Yes") * 100 #16.09195%
attrition_positive = filter(raw_data, Attrition == "No")
dim(attrition_positive)[1] #730, 83.9%
attrition_negative = filter(raw_data, Attrition == "Yes")
dim(attrition_negative)[1] #140, 16.1%
```

## convert categorical data to numeric
```{r}
raw_data$Attrition_num <- ifelse(raw_data$Attrition == "Yes", 1, 0)
```

## Explore the spread of all features
```{r}
raw_data %>%
  plot_boxplot(by = "Attrition")
```

## Look at attrition by job role, department, marital status
```{r}
raw_data %>%
  group_by(Department) %>%
  #group_by(JobRole) %>%
  #group_by(MaritalStatus) %>%
  summarize(n = n(),
            attrition_rate = mean(Attrition == "Yes") * 100
  ) %>%
  arrange(desc(attrition_rate)) %>%
  ggplot(aes(x = reorder(Department, 
  # ggplot(aes(x = reorder(JobRole, 
  # ggplot(aes(x = reorder(MaritalStatus, 
                         attrition_rate), 
           y = attrition_rate,
           fill = Department)) +
          # fill = JobRole)) +
          # fill = MaritalStatus)) +
  geom_col() +coord_flip() +
  geom_text(aes(label = 
                  paste0(round(attrition_rate, 1), "%")),
            vjust = 0,
            hjust = 1.1,
            size = 5,
            fontface = "bold") +
  labs(
    title = "Attrition Rate by Department",
    # title = "Attrition Rate by Job Role",
    # title = "Attrition Rate by Marital Status",
    x = NULL,
    y = "Attrition Rate (%)"
  ) +
  theme_minimal(base_size = 14)+
  theme(legend.position = "none")
```

```{r}
raw_data$OverTime_num <- ifelse(raw_data$OverTime == "Yes", 1, 0)
```

## Plot attrition percentage for job involvement, stock option level
```{r}
raw_data %>%
  # group_by(JobInvolvement) %>%
  group_by(StockOptionLevel) %>%
  summarise(
    n = n(),
    attrition_rate = mean(Attrition == "Yes") * 100
  ) %>%
  # arrange(JobInvolvement) %>%
  arrange(StockOptionLevel) %>%
  # ggplot(aes(x = factor(JobInvolvement),
  ggplot(aes(x = factor(StockOptionLevel),
           y = attrition_rate)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = 
                  paste0(round(attrition_rate, 1), "%")),
            vjust = 1,
            size = 5,
            fontface = "bold") +
  geom_hline(yintercept = overall_rate,
             linetype = "dashed",
             color = "red",
             size = 1) +
  labs(
    # title = "Attrition Rate by Job Involvement",
    title = "Attrition Rate by Stock Option Level",
    # x = "Job Involvement",
    x = "Stock Option Level",
    y = "Attrition Rate (%)"
  ) +
  theme_minimal(base_size = 14)
```
## Try the ggbetweenstats to visualize total working years
```{r}
# Violin plot, box plot, scatter plot, and density plot
library(ggstatsplot)
plt <- ggbetweenstats(
  data = raw_data,
  x = Attrition,
  y = TotalWorkingYears
)
plt
```

## Visualize attrition rate by overtime
```{r}
raw_data %>%
  group_by(OverTime) %>%
  summarise(
    n = n(),
    attrition_rate = mean(Attrition == "Yes") * 100
  ) %>%
  ggplot(aes(x = OverTime,
           y = attrition_rate,
           fill = OverTime)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(attrition_rate, 1), "%")),
            vjust = 1,
            size = 5,
            fontface = "bold") +
  geom_hline(yintercept = overall_rate,
             linetype = "dashed",
             color = "red",
             size = 1) +
  labs(
    title = "Attrition Rate by Overtime Status",
    subtitle = paste("Overall Attrition Rate:", 
                     round(overall_rate, 1), "%"),
    x = "Overtime",
    y = "Attrition Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("Yes" = "#e88631",
                               "No"  = "#7c3896")) + 
  theme(legend.position = "none")
```

## Correlation Heatmap, Bar Plot, GG Pairs of Selected Features
```{r}
raw_data %>% select(
  "Attrition",
  "JobInvolvement",
  # "JobLevel",
  # "StockOptionLevel",
  "MonthlyIncome",
  # "TotalWorkingYears",
  # "MaritalStatus",
  # "JobRole",
  "OverTime"
# ) %>% plot_correlation()
# ) %>% plot_bar()
) %>% ggpairs(aes(color = Attrition))
```

























## Classification Models
```{r}
library(class)
library(caret)
library(e1071) # Naive Bayes
```

### Change response variable to a factor and make "Yes" the positive class
```{r}
# get fresh raw data
raw_data <- read.csv("CaseStudy1-data.csv")
# make a copy of the raw_data for modeling
model_data <- raw_data %>% select(
                          "Attrition"
                          ,"Age"
                          ,"DistanceFromHome"
                          ,"JobInvolvement"
                          ,"JobLevel"
                          ,"JobSatisfaction"
                          ,"MaritalStatus"
                          ,"MonthlyIncome"
                          ,"NumCompaniesWorked"
                          ,"OverTime"
                          ,"WorkLifeBalance"
                          ,"YearsAtCompany"
                          ,"YearsSinceLastPromotion"
                          )
```
## Train/Test Split
```{r}
#split and seed can be changed
set.seed(60)
split = .67

obs = dim(model_data)[1]
train_inds_knn = sample(seq(1:obs),round(split*obs))
training_set_knn = model_data[train_inds_knn,]
testing_set_knn = model_data[-train_inds_knn,]
```

## Feature Engineering
"Data leakage":
You should compute engineered features using only the training data 
so the model does not accidentally learn information from the test set.

```{r}
# Split monthly income into four levels based on quartile
income_breaks <- quantile(training_set_knn$MonthlyIncome,
                          probs = c(0, 0.25, 0.5, 0.75, 1))
training_set_knn$income_quartile_labels <- cut(training_set_knn$MonthlyIncome,
                               breaks = income_breaks,
                               include.lowest = TRUE,
                               labels = c("Q1_Low",
                                          "Q2_MidLow",
                                          "Q3_MidHigh",
                                          "Q4_High"))
training_set_knn$income_quartile_num <- as.numeric(training_set_knn$income_quartile_labels)
testing_set_knn$income_quartile_labels <- cut(testing_set_knn$MonthlyIncome,
                               breaks = income_breaks,
                               include.lowest = TRUE,
                               labels = c("Q1_Low",
                                          "Q2_MidLow",
                                          "Q3_MidHigh",
                                          "Q4_High"))
testing_set_knn$income_quartile_num <- as.numeric(testing_set_knn$income_quartile_labels)

# look at median income
median_attrition_income <- median(
                        training_set_knn$MonthlyIncome[training_set_knn$Attrition == "Yes"]
                      )
training_set_knn$low_income_risk <- ifelse(
  training_set_knn$MonthlyIncome < median_attrition_income, 1, 0
)
testing_set_knn$low_income_risk <- ifelse(
  testing_set_knn$MonthlyIncome < median_attrition_income, 1, 0
)

# marital status to marital attrition risk
training_set_knn$marital_status_single <- ifelse(training_set_knn$MaritalStatus == "Single", 1, 0)
training_set_knn$marital_attrition_risk <- dplyr::case_when(
                                    training_set_knn$MaritalStatus == "Single" ~ 2,
                                    training_set_knn$MaritalStatus == "Married" ~ 1,
                                    training_set_knn$MaritalStatus == "Divorced" ~ 0
                                  )

testing_set_knn$marital_status_single <- ifelse(testing_set_knn$MaritalStatus == "Single", 1, 0)
testing_set_knn$marital_attrition_risk <- dplyr::case_when(
                                    testing_set_knn$MaritalStatus == "Single" ~ 2,
                                    testing_set_knn$MaritalStatus == "Married" ~ 1,
                                    testing_set_knn$MaritalStatus == "Divorced" ~ 0
                                  )

# Burnout Index: Overtime + (4- Work Life Balance) + (4- Job Satisfaction)
training_set_knn$OverTime_num <- ifelse(training_set_knn$OverTime == "Yes", 1, 0)
training_set_knn$burnout_index <-
  training_set_knn$OverTime_num + 
  (4-training_set_knn$WorkLifeBalance) + 
  (4-training_set_knn$JobSatisfaction)
testing_set_knn$OverTime_num <- ifelse(testing_set_knn$OverTime == "Yes", 1, 0)
testing_set_knn$burnout_index <-
  testing_set_knn$OverTime_num + 
  (4-testing_set_knn$WorkLifeBalance) + 
  (4-testing_set_knn$JobSatisfaction)

# Pay Frustration: (Job Level + Job Involvement) / Monthly Income
training_set_knn$pay_frustration <-
  (training_set_knn$JobLevel + training_set_knn$JobInvolvement)*2000 /
  training_set_knn$income_quartile_num
testing_set_knn$pay_frustration <-
  (testing_set_knn$JobLevel + testing_set_knn$JobInvolvement)*2000 /
  testing_set_knn$income_quartile_num

# Career Stagnation: Years Since Last Promotion / (Years at Company +1)
training_set_knn$career_stagnation <-
  training_set_knn$YearsSinceLastPromotion /
  (training_set_knn$YearsAtCompany + 1)
testing_set_knn$career_stagnation <-
  testing_set_knn$YearsSinceLastPromotion /
  (testing_set_knn$YearsAtCompany + 1)

# Commute Pain: Distance from Home * Overtime
training_set_knn$commute_pain <-
  training_set_knn$DistanceFromHome *
  training_set_knn$OverTime_num
testing_set_knn$commute_pain <-
  testing_set_knn$DistanceFromHome *
  testing_set_knn$OverTime_num

# job hopping rate based on marital status, num companies, age
training_set_knn$job_hopping_rate <-
  training_set_knn$marital_attrition_risk *
  (training_set_knn$NumCompaniesWorked / 
  training_set_knn$Age)
testing_set_knn$job_hopping_rate <-
  testing_set_knn$marital_attrition_risk *
  (testing_set_knn$NumCompaniesWorked / 
  testing_set_knn$Age)

# compensation unfairness based on income, job involvement, overtime
training_set_knn$compensation_unfairness <- 
  training_set_knn$income_quartile_num * 
  training_set_knn$JobInvolvement *
  (1+training_set_knn$OverTime_num)
testing_set_knn$compensation_unfairness <- 
  testing_set_knn$income_quartile_num * 
  testing_set_knn$JobInvolvement *
  (1+testing_set_knn$OverTime_num)

# reorder the levels so Attrition Yes is the positive class
training_set_knn$Attrition <- as.factor(training_set_knn$Attrition)
levels(training_set_knn$Attrition)
training_set_knn$Attrition <- relevel(training_set_knn$Attrition, "Yes")
levels(training_set_knn$Attrition)

testing_set_knn$Attrition <- factor(testing_set_knn$Attrition,
                                    levels = levels(training_set_knn$Attrition))

#scaling
training_set_knn_scaled <- scale(training_set_knn[,c(
  "burnout_index"
  ,"career_stagnation"
  ,"pay_frustration"
  ,"job_hopping_rate"
  ,"compensation_unfairness"
  ,"commute_pain"
)])

testing_set_knn_scaled <- scale(
  testing_set_knn[,c(
    "burnout_index"
    ,"career_stagnation"
    ,"pay_frustration"    
    ,"job_hopping_rate"
    ,"compensation_unfairness"
    ,"commute_pain"
  )],
  center = attr(training_set_knn_scaled,"scaled:center"),
  scale  = attr(training_set_knn_scaled,"scaled:scale")
)

```


## KNN model (not optimized)
```{r}
# tunable threshold, hyperparameter k
classification_threshold = 0.104
k_value = 19
```


```{r}
set.seed(47)
preds <- knn(training_set_knn_scaled,
             testing_set_knn_scaled,
             training_set_knn$Attrition,
             k = k_value, 
             prob = TRUE)

# use the threshold
prob <- attr(preds, "prob")
prob_yes <- ifelse(preds == "Yes", prob, 1 - prob)
preds2 <- ifelse(prob_yes > classification_threshold, "Yes", "No")
preds2 <- factor(preds2, levels = levels(testing_set_knn$Attrition))

# confusion matrix
cm_knn <- confusionMatrix(preds2, testing_set_knn$Attrition)
cm_knn$table

accuracy_knn_1 <- cm_knn$overall[1] #accuracy
sensitivity_knn_1 <- cm_knn$byClass[1] #sensitivity
specificity_knn_1 <- cm_knn$byClass[2] #specificity
accuracy_knn_1
sensitivity_knn_1
specificity_knn_1
```


## Tune the hyperparameter k
We find that k=19 has the highest sensitivity. 

```{r}
# I only want to pick an odd numbered k
set.seed(60)
numks = 120 # For each of the 60 values of k that we are trying
iterations = 30 # obtain an accuracy statistic for 30 different train test splits
split = .7
obs = dim(raw_data)[1]

# I only want the highest sensitivity
masterSens = matrix(nrow = iterations, ncol = numks)

for(j in 1:iterations)
{

  trainIndices = sample(seq(1:obs),round(split*obs))
  trainSet = raw_data[trainIndices,]
  testSet = raw_data[-trainIndices,]
  
  train_knn <- scale(trainSet[,c(
                              "burnout_index"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              )])

  test_knn  <- scale(testSet[,c(
                              "burnout_index"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              )],
                   center = attr(train_knn,"scaled:center"),
                   scale  = attr(train_knn,"scaled:scale"))
  for(i in seq(1, numks, by = 2))
    {
      preds <- knn(train_knn,
               test_knn,trainSet$Attrition,
               k = i, 
               prob = TRUE)
  
      prob <- attr(preds, "prob")
      prob_yes <- ifelse(preds == "Yes", prob, 1 - prob)
      preds2 <- ifelse(prob_yes > classification_threshold, "Yes", "No")
      preds2 <- factor(preds2, levels = levels(testSet$Attrition))
      
      cm <- preds2 %>%
        table(testSet$Attrition) %>%
        confusionMatrix()
      masterSens[j,i] = cm$byClass[1]
    }
}

MeanSens = colMeans(masterSens)
plot(seq(1,numks,1),MeanSens, type = "l", ylim = c(.5,1))
which.max(MeanSens) # which k value has the highest sensitivity? k=13
max(MeanSens)
```


## Tune the classification threshold
```{r}
preds <- knn(train_knn,
             test_knn,trainSet$Attrition,
             k = 19, 
             prob = TRUE)
prob <- attr(preds, "prob")
prob_yes <- ifelse(preds == "Yes", prob, 1 - prob)

# zero in on the threshold range:
# thresholds <- seq(0.45, 0.05, by = -0.05)
# thresholds <- seq(0.3, 0.1, by = -0.01)
thresholds <- seq(0.11, 0.09, by = -0.001)
length(thresholds)

accs = data.frame(
  accuracy = numeric(length(thresholds)),
  sensitivity = numeric(length(thresholds)),
  specificity = numeric(length(thresholds)),
  yes_threshold = thresholds
)

for (j in seq_along(thresholds)) {

  t <- thresholds[j]

  preds2 <- ifelse(prob_yes > t, "Yes", "No")
  preds2 <- factor(preds2, levels = levels(testSet$Attrition))

  cm <- preds2 %>%
    table(testSet$Attrition) %>%
    confusionMatrix()

  accs$accuracy[j] <- cm$overall[1]
  accs$sensitivity[j] <- cm$byClass[1]
  accs$specificity[j] <- cm$byClass[2]
}

max_sens <- max(accs$sensitivity)
best_sens <- accs$sensitivity[accs$sensitivity == max_sens]
best_acc <-accs$accuracy[accs$sensitivity == max_sens]
best_spec<-accs$specificity[accs$sensitivity == max_sens] 

#get the highest sensitivity where specificity is at least 0.6
valid <- accs %>%
  filter(specificity >= 0.6)
best_row <- valid %>%
  slice_max(sensitivity, n = 1)
best_threshold <- max(best_row$yes_threshold)
best_sens <- max(best_row$sensitivity)
best_spec <- max(best_row$specificity)
best_acc <- max(best_row$accuracy)

best_threshold #0.104
best_sens #0.69
best_spec #0.68
best_acc #0.68
```

## Plot the threshold accuracies
```{r}
library(tidyr)
library(ggplot2)

accs_long <- accs |>
  pivot_longer(
    cols = c(sensitivity, specificity),
    names_to = "metric",
    values_to = "value"
  )

label_text <- paste0(
  "Threshold = ", round(best_threshold, 2),
  "\nSensitivity = ", round(best_sens, 2),
  "\nSpecificity = ", round(best_spec, 2)
)

ggplot(accs_long,
       aes(x = yes_threshold,
           y = value,
           color = metric)) +

  geom_line(size = 1.3) +

  geom_point(size = 3) +

  geom_hline(yintercept = 0.6,
             linetype = "dashed",
             color = "gray40") +
  geom_vline(xintercept = best_threshold,
           linetype = "dashed",
           color = "darkgreen",
           size = 1) +
  
    annotate("text",
           x = best_threshold + 0.02,
           y = 0.8,
           label = label_text,
           hjust = 0,
           size = 4,
           color = "black") +

  scale_color_manual(values = c("steelblue","firebrick")) +

  labs(
    title = "Effect of Classification Threshold on Model Performance",
    x = "Probability Threshold for Predicting Attrition = Yes",
    y = "Model Performance",
    color = "Metric"
  ) +

  theme_minimal(base_size = 14)

```

## Test the optimized KNN model
```{r}
set.seed(60)
split = .7

obs = dim(raw_data)[1]
trainIndices = sample(seq(1:obs),round(split*obs))
trainSet = raw_data[trainIndices,]
testSet = raw_data[-trainIndices,]
classification_threshold = 0.104
k_value = 19
train_knn <- scale(trainSet[,c(
                              "burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                              ,"compensation_unfairness"
                              )])

test_knn  <- scale(testSet[,c(
                              "burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                              ,"compensation_unfairness"
                              )],
                   center = attr(train_knn,"scaled:center"),
                   scale  = attr(train_knn,"scaled:scale"))
preds <- knn(train_knn,
             test_knn,trainSet$Attrition,
             k = k_value, 
             prob = TRUE)

prob <- attr(preds, "prob")

prob_yes <- ifelse(preds == "Yes", prob, 1 - prob)

preds2 <- ifelse(prob_yes > classification_threshold, "Yes", "No")
preds2 <- factor(preds2, levels = levels(testSet$Attrition))

cm <- preds2 %>%
  table(testSet$Attrition) %>%
  confusionMatrix()
cm
```

## Naive Bayes Model
```{r}
model <- naiveBayes(trainSet[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                              # ,"compensation_unfairness"
                                 )],
                    trainSet$Attrition)

pred_probs <- predict(model,
                 testSet[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                              # ,"compensation_unfairness"
                             )],
                 type = "raw")
# 
# 
# pred_class <- ifelse(pred_probs[, "Yes"] > 0.1, "Yes", "No")
# CM_NB <- confusionMatrix(factor(pred_class, levels=c("No","Yes")),
#                 testSet$Attrition)
# CM_NB$overall[1]
# CM_NB$byClass[1]
# CM_NB$byClass[2]
```
```{r}
prob_yes <- pred_probs[, "Yes"]

# zero in on the threshold range
# thresholds <- seq(0.5, 0.01, by = -0.01)
# thresholds <- seq(0.3, 0.05, by = -0.01)
thresholds <- seq(0.1, 0.01, by = -0.001)
# length(thresholds)

accs_nb <- data.frame(
  accuracy = numeric(length(thresholds)),
  sensitivity = numeric(length(thresholds)),
  specificity = numeric(length(thresholds)),
  yes_threshold = thresholds
)

for (j in seq_along(thresholds)) {

  t <- thresholds[j]

  preds <- ifelse(prob_yes > t, "Yes", "No")
  preds <- factor(preds, levels = levels(testSet$Attrition))

  cm <- confusionMatrix(preds, testSet$Attrition)

  accs_nb$accuracy[j] <- cm$overall[1]
  accs_nb$sensitivity[j] <- cm$byClass[1]
  accs_nb$specificity[j] <- cm$byClass[2]
}

valid_nb <- accs_nb %>%
  filter(specificity >= 0.6)

best_nb <- valid_nb %>%
  slice_max(sensitivity, n = 1)

best_nb
```

## Visualize the comparison between models
```{r}
accs$model <- "KNN"
accs_nb$model <- "Naive Bayes"
all_accs <- rbind(accs, accs_nb)

all_long <- all_accs |>
  pivot_longer(
    cols = c(sensitivity, specificity),
    names_to = "metric",
    values_to = "value"
  )
ggplot(all_long,
       aes(x = yes_threshold,
           y = value,
           color = metric)) +

  geom_line(size = 1.2) +
  geom_point(size = 2) +
  
  geom_hline(yintercept = 0.6,
             linetype = "dashed",
             color = "gray40") +

  facet_wrap(~model) +

  scale_color_manual(values = c("steelblue","firebrick")) +

  labs(
    title = "Threshold Tuning Comparison: k-NN vs Naive Bayes",
    x = "Probability Threshold for Predicting Attrition = Yes",
    y = "Model Performance",
    color = "Metric"
  ) +

  theme_minimal(base_size = 14)
```

## Test optimized NB model
```{r}
model <- naiveBayes(trainSet[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                                 )],
                    trainSet$Attrition)

pred_probs <- predict(model,
                 testSet[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                             )],
                 type = "raw")


pred_class <- ifelse(pred_probs[, "Yes"] > 0.066, "Yes", "No")
CM_NB <- confusionMatrix(factor(pred_class, levels=c("No","Yes")),
                testSet$Attrition)
CM_NB
```

```{r}

```






