---
title: "Comparing Classification Based on Male and Female Titanic Passengers"
author: "Isabella Westlauer"
date: "2026-02-07"
output: html_document
---

## Get the titanic data using a GitHub link

Fetch the data, inspect it and tidy it.

```{r}
library(class)
library(caret)
library(e1071)
library(tidyverse)
library(plotly)
library(patchwork)
```

```{r}
url <- "https://raw.githubusercontent.com/BivinSadler/MSDS_6306_Doing-Data-Science/Master/Unit%206/titanic_train.csv"
titanic <- read.csv(url)
titanic <- titanic[!is.na(titanic$Age), ] #remove observations with NA Age
titanic$Survived <- factor(titanic$Survived, labels = c("Died", "Survived")) #0 is survived??
```

Create split datasets for males and females
```{r}
titanic_males <- filter(titanic, titanic$Sex == "male") # how can I use regex here to get male, Male, M?
titanic_females <- filter(titanic, titanic$Sex == "female")
# titanic_NA_Sex <- filter(titanic, is.na(titanic$Sex)) # there are no NA Sex?
# 
# head(titanic_females)
# # how many females actually died? -- 197 or 75%
# count(filter(titanic_females, titanic_females$Survived == "Died"))
```


## Partition the data  

Start with females
```{r}
# dim(titanic_females) #not the same number of observations as titanic
# dim(titanic_males)
# dim(titanic_NA_Sex) # no na values in titanic$Sex

# the number of males and females in the dataset is 714, not 891?
set.seed(6)
rows_f = dim(titanic_females)[1] # number of rows is 261
splitPerc = (600/891) # put 176 out of 261 in the training set
trainIndices_titanic_f = sample(1:rows_f,round(splitPerc * rows_f))
trn_f = titanic_females[trainIndices_titanic_f,]
tst_f = titanic_females[-trainIndices_titanic_f,]
length(tst_f) #there are only 14 observations in the test set?... no, 14 columns
dim(tst_f)[1] #85 observations in the test set

set.seed(6)
rows_m = dim(titanic_males)[1] # number of rows is 453
splitPerc = (600/891) # put 176 out of 261 in the training set
trainIndices_titanic_m = sample(1:rows_m,round(splitPerc * rows_m))
trn_m = titanic_males[trainIndices_titanic_m,]
tst_m = titanic_males[-trainIndices_titanic_m,]
dim(tst_m)[1] #148 observations in the test set
```

## Female Passenger Classifications with K = 3, 5, 10, 20  

```{r}
# k = 3
classifications_f = knn(trn_f[,c(3,6)],
                      tst_f[,c(3,6)],
                      trn_f$Survived, 
                      prob = TRUE, k = 3)
confusionMatrix(table(classifications_f,tst_f$Survived))
# k = 5
classifications_f = knn(trn_f[,c(3,6)],
                      tst_f[,c(3,6)],
                      trn_f$Survived, 
                      prob = TRUE, k = 5)
confusionMatrix(table(tst_f$Survived,classifications_f))
# k = 10
classifications_f = knn(trn_f[,c(3,6)],
                      tst_f[,c(3,6)],
                      trn_f$Survived, 
                      prob = TRUE, k = 10)
confusionMatrix(table(tst_f$Survived,classifications_f))
# k = 20
classifications_f = knn(trn_f[,c(3,7)],
                      tst_f[,c(3,7)],
                      trn_f$Survived, 
                      prob = TRUE, k = 20)
confusionMatrix(table(tst_f$Survived,classifications_f))
```
```{r}

# k = 3
classifications_m = knn(trn_m[,c(3,6)],
                      tst_m[,c(3,6)],
                      trn_m$Survived, 
                      prob = TRUE, k = 3)
confusionMatrix(table(classifications_m,tst_m$Survived))
# k = 5
classifications_m = knn(trn_m[,c(3,6)],
                      tst_m[,c(3,6)],
                      trn_m$Survived, 
                      prob = TRUE, k = 5)
confusionMatrix(table(tst_m$Survived,classifications_m))
# k = 10
classifications_m = knn(trn_m[,c(3,6)],
                      tst_m[,c(3,6)],
                      trn_m$Survived, 
                      prob = TRUE, k = 10)
confusionMatrix(table(tst_m$Survived,classifications_m))
# k = 20
classifications_m = knn(trn_m[,c(3,7)],
                      tst_m[,c(3,7)],
                      trn_m$Survived, 
                      prob = TRUE, k = 20)
confusionMatrix(table(tst_m$Survived,classifications_m))
```


## Loop for many k values  

```{r}
accs = data.frame(accuracy = numeric(30), k = numeric(30))

for(i in 1:30)
{
  classifications = knn(trn[,c(3,6)],tst[,c(3,6)],trn$Survived, prob = TRUE, k = i)
  table(tst$Survived,classifications)
  CM = confusionMatrix(table(tst$Survived,classifications))
  accs$accuracy[i] = CM$overall[1]
  accs$k[i] = i
}

plot(accs$k,accs$accuracy, type = "l", xlab = "k")

```
## Loop for many k and the average of many training / test partitions  
```{r}
iterations = 500
numks = 30

masterAcc = matrix(nrow = iterations, ncol = numks)
  
for(j in 1:iterations)
{
accs = data.frame(accuracy = numeric(30), k = numeric(30))
trainIndices = sample(1:dim(titanic)[1],600)
trn = titanic[trainIndices,]
tst = titanic[-trainIndices,]
for(i in 1:numks)
{
  classifications = knn(trn[,c(3,6)],
                        tst[,c(3,6)],
                        trn$Survived, 
                        prob = TRUE, k = i)
  table(classifications,tst$Survived)
  CM = confusionMatrix(table(classifications,tst$Survived))
  masterAcc[j,i] = CM$overall[1]
}
}

MeanAcc = colMeans(masterAcc) # mean accuracy versus different k values
plot(seq(1,numks,1),MeanAcc, type = "l") # how do I use plotly to get interactive?
```
Pick an age and predict the survival probability of a person that age based on each of the ticket classes.  

```{r}

train_df <- trn %>%
  select(Survived, Pclass, Age)
```
```{r}
test_df = data.frame(Age = 75.0, Pclass = 1)
knn(train_df[,2:3], test_df, train_df$Survived, k = 10, prob = TRUE) 
```
```{r}
test_df = data.frame(Age = 75.0, Pclass = 2)
knn(train_df[,2:3], test_df, train_df$Survived, k = 10, prob = TRUE) 
```
```{r}
test_df = data.frame(Age = 75.0, Pclass = 3)
knn(train_df[,2:3], test_df, train_df$Survived, k = 10, prob = TRUE) 
```

How do I generate a test_df that has many combination of age and class?
```{r}
# actually... after I do that I want to use standardization on Age and Class,
# because I think a very young or very old age is significant when looking at survival

# 1. Initialize an empty data frame with column names and types
test_df <- data.frame(Pclass = numeric(0), Age = numeric(0))
# 2. Use a for loop to iterate from 1 to 5
for (i in 1:3) { #Add classes from 1 to 3
  for (j in 0:75) { #Add ages from 0 to 75
    new_age <- j
    new_pclass <- i
    new_row<- data.frame(Age = new_age, Pclass = new_pclass)
    test_df<- rbind(test_df, new_row)
  }
}

# I don't need a nested for loop to do this, do I?
# test_df_better = c(seq(1,3,1), seq(0,75,1))
```

KNN Cross Validation
```{r}
# Choose k=10 because of the previous graph
knn(train_df[,2:3], test_df, train_df$Survived, k = 10, prob = TRUE)
# with cross-validation
classifications = knn.cv(train_df[,2:3],train_df$Survived, k = 10)
# confusionMatrix expects the same levels in the same order
confusionMatrix(classifications,train_df$Survived)

```
