---
title: "Unit 6 ML Classification and KNN"
author: "Isabella Westlauer"
date: "2026-02-03"
output: html_document
---

## Get the titanic data using a GitHub link

Fetch the data, inspect it and tidy it.

```{r}
#install package and load library
# install.packages("class")
# install.packages("caret")
# install.packages("e1071")
# install.packages("tidyverse")  # only needs to be run once

library(class)
library(caret)
library(e1071)
library(tidyverse)
library(plotly)
library(ggthemes)


# Assign the raw GitHub URL to a variable
url <- "https://raw.githubusercontent.com/BivinSadler/MSDS_6306_Doing-Data-Science/Master/Unit%206/titanic_train.csv"

# Read the data into a data frame
titanic <- read.csv(url)

# View the first few rows of the data
head(titanic)
```
```{r}

# Question -- what is a better way to find the index of a column if you know the header?

# column 1 is PassengerID, column2 is Survived
# Class is col3, Age is col6

sum(is.na(titanic)) #177 NA values in the dataset
sum(is.na(titanic$Age))
titanic <- titanic[!is.na(titanic$Age), ]

sum(is.na(titanic)) # got rid of all the NA values!

# put survived in as "Survived" and "Died"
titanic$Survived <- factor(titanic$Survived, labels = c("Died", "Survived"))

```
## Partition the data  

```{r}
set.seed(6) # What does this do? Guarantees the same random values, makes this reproducible.
splitPerc = 600/891
# dim(titanic) is 891 rows, 12 columns
rows = dim(titanic)[1] # number of rows in titanic data = 891
trainIndices_titanic = sample(1:rows,round(splitPerc * rows)) # sample(1:891, 600)
# round(splitPerc * rows) # 600 in the training set
trn = titanic[trainIndices_titanic,]
tst = titanic[-trainIndices_titanic,]
# dim(test) #291 rows
```

## Classifications with K = 3, 5, 10, 20  

```{r}
# k = 3
classifications = knn(trn[,c(3,6)],
                      tst[,c(3,6)],
                      trn$Survived, 
                      prob = TRUE, k = 3)
# table(classifications,tst$Survived)
confusionMatrix(table(classifications,tst$Survived))
```
```{r}
# k = 5
classifications = knn(trn[,c(3,6)],
                      tst[,c(3,6)],
                      trn$Survived, 
                      prob = TRUE, k = 5)
# table(tst$Survived,classifications)
confusionMatrix(table(tst$Survived,classifications))
```
```{r}
# k = 10
classifications = knn(trn[,c(3,6)],
                      tst[,c(3,6)],
                      trn$Survived, 
                      prob = TRUE, k = 10)
# table(tst$Survived,classifications)
confusionMatrix(table(tst$Survived,classifications))
```
```{r}
# k = 20
classifications = knn(trn[,c(3,7)],tst[,c(3,7)],
                      trn$Survived, prob = TRUE, k = 20)
# table(tst$Survived,classifications)
confusionMatrix(table(tst$Survived,classifications))
# CM = confusionMatrix(table(tst$Survived,classifications))
# CM$overall[1] # Accuracy 0.6907216

```

## Loop for many k values  

```{r}
accs = data.frame(accuracy = numeric(30), k = numeric(30))

for(i in 1:30)
{
  classifications = knn(trn[,c(3,6)],tst[,c(3,6)],trn$Survived, prob = TRUE, k = i)
  table(tst$Survived,classifications)
  CM = confusionMatrix(table(tst$Survived,classifications))
  accs$accuracy[i] = CM$overall[1]
  accs$k[i] = i
}

max_acc <- max(accs$accuracy)
best_k  <- accs$k[accs$accuracy == max_acc]
best_acc <- accs$accuracy[accs$accuracy == max_acc]
```
## Plots

```{r}
plot(accs$k, accs$accuracy,
     type = "l",
     xlab = "k",
     main = "k value versus accuracy for k=1-30")

points(best_k, best_acc,
       pch = 19,
       col = "red",
       cex = 1.5)

abline(v = best_k,
       col = "red",
       lty = 2,
       lwd = 2)

par(xpd = TRUE)
text(best_k, best_acc,
     labels = paste("k =", best_k),
     pos = 3,
     offset = 0.75,
     col = "red")
par(xpd = FALSE)


```
## Loop for many k and the average of many training / test partitions  
```{r}
iterations = 500
numks = 90

masterAcc = matrix(nrow = iterations, ncol = numks)
  
for(j in 1:iterations)
{
accs = data.frame(accuracy = numeric(90), k = numeric(90))
trainIndices = sample(1:dim(titanic)[1],600)
trn = titanic[trainIndices,]
tst = titanic[-trainIndices,]
for(i in 1:numks)
{
  classifications = knn(trn[,c(3,6)],
                        tst[,c(3,6)],
                        trn$Survived, 
                        prob = TRUE, k = i)
  table(classifications,tst$Survived)
  CM = confusionMatrix(table(classifications,tst$Survived))
  masterAcc[j,i] = CM$overall[1] # this refers to the accuracies
}
}

MeanAcc = colMeans(masterAcc) # mean accuracy versus different k values
```

```{r}
plot(seq(1,numks,1),MeanAcc, 
     type = "l",
     xlab = "k",
     ylab = "mean accuracy",
     main = "master accuracy of k values 1 to 30")

max_mean_acc <- max(MeanAcc)
best_k <- which(MeanAcc == max_mean_acc)

points(best_k, MeanAcc[best_k],
       pch = 19,
       col = "red")

abline(h = max_mean_acc,
       col = "red",
       lty = 2,
       lwd = 2)

abline(v = best_k,
       col = "red",
       lty = 2,
       lwd = 2)
par(xpd = TRUE)
text(best_k, max_mean_acc,
     labels = paste("k =", best_k),
     pos = 3,
     offset = 0.75,
     col = "red")
par(xpd = FALSE)

```
Pick an age and predict the survival probability of a person that age based on each of the ticket classes.  

```{r}
# choose only survived, class, and age from the training dataset
# train_df <- trn %>% select(Survived, Pclass, Age)
# str(train_df) 
# str(classifications)
# okay Survived is an int but needs to be a factor

train_df <- trn %>%
  select(Survived, Pclass, Age) %>%
  mutate(Survived = factor(Survived, levels = c(0, 1)))

# What are all the levels of Pclass? ha! it works just like Excel!
# unique(titanic$Pclass)
# try one data point, age 75 and class 3
# Had to open the csv and use Excel fmlas to check this
```
```{r}
test_df = data.frame(Age = 75, Pclass = 1)
knn(train_df[,2:3], test_df, train_df$Survived, k = 5, prob = TRUE) # 75% survived
```
```{r}
test_df = data.frame(Age = 75, Pclass = 2)
knn(train_df[,2:3], test_df, train_df$Survived, k = 5, prob = TRUE) # 80% died
```
```{r}
test_df = data.frame(Age = 75, Pclass = 3)
knn(train_df[,2:3], test_df, train_df$Survived, k = 5, prob = TRUE) # 60% died
```

How do I generate a test_df that has many combination of age and class?
```{r}
# actually... after I do that I want to use standardization on Age and Class,
# because I think a very young or very old age is significant when looking at survival

# 1. Initialize an empty data frame with column names and types
test_df <- data.frame(Pclass = numeric(0), Age = numeric(0))
# 2. Use a for loop to iterate from 1 to 5
for (i in 1:3) { #Add classes from 1 to 3
  for (j in 0:75) { #Add ages from 0 to 75
    new_age <- j
    new_pclass <- i
    new_row<- data.frame(Age = new_age, Pclass = new_pclass)
    test_df<- rbind(test_df, new_row)
  }
}

# I don't need a nested for loop to do this, do I?
# test_df_better = c(seq(1,3,1), seq(0,75,1))
```

KNN Cross Validation
```{r}
# Choose k=10 because of the previous graph
knn(train_df[,2:3], test_df, train_df$Survived, k = 10, prob = TRUE)
# with cross-validation
classifications = knn.cv(train_df[,2:3],train_df$Survived, k = 10)
# confusionMatrix expects the same levels in the same order
confusionMatrix(classifications,train_df$Survived)

```

Since I want to consider a passenger age 75, I need to standardize.
```{r}
#create standardized variables
titanic$Z_Pclass = scale(titanic$Pclass)
titanic$Z_Age = scale(titanic$Age)

#Internal CV
#Raw Limit and AGE
classifications = knn.cv(train_df[,2:3],train_df$Survived, prob=TRUE, k = 10)
confusionMatrix(classifications,train_df$Survived)

#make titanic's survived into levels
titanic_leveled <- titanic %>%
  select(Survived, Pclass, Age, Z_Pclass, Z_Age) %>%
  mutate(Survived = factor(Survived, levels = c(0, 1)))

#Standardized
classifications = knn.cv(titanic_leveled[,4:5],
                         titanic_leveled$Survived, 
                         prob=TRUE, 
                         k = 10)
confusionMatrix(classifications,
                titanic_leveled$Survived)

titanic_leveled$classifications<- classifications

titanic_leveled %>% 
  ggplot(aes(x = Age,Pclass,color = classifications)) + 
  geom_point() +
  ggtitle("Passenger Age and Class by Classification")
```
