---
title: "iris_multinomial"
author: "Isabella Westlauer"
date: "2026-02-06"
output: html_document
---

For the full (multinomial) IRIS data (the iris dataset in R), 
do a 70-30 train/test cross validation with 
k =1 - 90 and use sepal length and width as predictors.  
Make a plot of k (xaxis) versus accuracy.  
Use this plot to tune the hyperparameter k.  
What do you feel is the best value of k?  
```{r}
library(tidyverse)
library(plotly)
library(class)
library(caret)
library(e1071)
library(ggthemes)
```

## Partition the Data

```{r}
set.seed(10) # for reproducibility
split_percent = 0.7
num_observations <- dim(iris)[1]
training_indices = sample(1:num_observations,round(split_percent * num_observations))
iris_training = iris[training_indices,]
iris_test = iris[-training_indices,]

```

## Plot just for fun

Sepal length and width are the explanatory variables, or predictors.
```{r}
iris %>% ggplot(aes(x = Sepal.Length,Sepal.Width,color = Species)) + 
  geom_point() +
  ggtitle("Iris sepal length and width according to species")
# Question-- if we think about a test with sepal.width >4.5, or sepal.length>8,
# Then we're extrapolating? We should not use knn classification? That test point
# would be an outlier and it would be strange to classify it based on the nearest
# neighbors which are far out.
```

Make a plot of k versus accuracy
```{r}
accuracies = data.frame(accuracy = numeric(90), 
                        k = numeric(90))

for(i in 1:90)
{
  classifications = knn(iris_training[,c(1,2)],
                        iris_test[,c(1,2)],
                        iris_training$Species, 
                        prob = TRUE, k = i)
  CM = confusionMatrix(table(iris_test$Species,
                             classifications))
  accuracies$accuracy[i] = CM$overall[1]
  accuracies$k[i] = i
}

p<- ggplot(accuracies, aes(k, accuracy)) + 
  geom_line() +
  ggtitle("Accuracy versus k value")

ggplotly(p + theme_economist())
```

```{r}
iterations = 500
numks = 90

masterAcc = matrix(nrow = iterations, ncol = numks)
  
for(j in 1:iterations)
{
accuracies = data.frame(accuracy = numeric(numks), 
                        k = numeric(numks))
training_indices = sample(1:dim(iris)[1],
                          0.7*dim(iris)[1])
iris_training = iris[training_indices,]
iris_test = iris[-training_indices,]
for(i in 1:numks)
{
  classifications = knn(iris_training[,c(1,2)],
                        iris_test[,c(1,2)],
                        iris_training$Species, 
                        prob = TRUE, k = i)
  table(classifications,iris_test$Species)
  CM = confusionMatrix(table(classifications,
                             iris_test$Species))
  masterAcc[j,i] = CM$overall[1]
}
}
# mean accuracy versus different k values
MeanAcc = colMeans(masterAcc) 
mean(MeanAcc)
```
```{r}
p <- plot(seq(1,numks,1),
          MeanAcc, 
          type = "l",
          xlab = "k",
          ylab = "Mean Accuracy",
          main = "Master Accuracy of k 1 to 90") # how do I use plotly to get interactive?


max_mean_acc <- max(MeanAcc)
best_k <- which(MeanAcc == max_mean_acc)

points(best_k, MeanAcc[best_k],
       pch = 19,
       col = "red")

abline(h = max_mean_acc,
       col = "red",
       lty = 2,
       lwd = 2)

abline(v = best_k,
       col = "red",
       lty = 2,
       lwd = 2)
par(xpd = TRUE)
text(best_k, max_mean_acc,
     labels = paste("k =", best_k),
     pos = 3,
     offset = 1,
     col = "red")
par(xpd = FALSE)




```

```{r}
```

```{r}
```

```{r}
```

```{r}
```