---
title: "Unit 5 For Live Session Solution"
author: "Bivin"
output: html_document
editor_options: 
  chunk_output_type: console
---


# Unit 5 Live Session R Code

## Part 1 BBall

Creating the TotalInches column:
```{r}
library(tidyverse)
BBall = read.csv(file.choose(),header = TRUE)
head(BBall)
BBall = BBall %>% separate(height, c("feet","inches"), remove = FALSE) %>% mutate(TotalInches = 12*as.numeric(feet) + as.numeric(inches))
head(BBall)


BBall[2143,]
#https://en.wikipedia.org/wiki/George_Karl
BBall[2143,"height"] = "6-2"
BBall[2143,"weight"] = "185"
BBall[2143,"TotalInches"] = 6*12 + 2
BBall[2143,"position"] = "G"

#Check our cleaning
BBall[2143,]


BBall = BBall %>% separate(height, c("feet","inches"), remove = FALSE) %>% mutate(TotalInches = 12*as.numeric(feet) + as.numeric(inches))
head(BBall)
class(BBall$TotalInches)

#note remove is now default to TRUE
BBall = BBall %>% separate(height, c("feet","inches")) %>% mutate(TotalInches = 12*as.numeric(feet) + as.numeric(inches))
head(BBall)

#Clean Weight
class(BBall$weight)
BBall$weight = as.numeric(BBall$weight)
class(BBall$weight)
```

Making the histogram:
```{r}
BBall %>% ggplot(aes(x = TotalInches, fill = position)) + geom_histogram()


BBall %>% ggplot(aes(x = TotalInches, color = position)) + geom_histogram() + facet_wrap(~position)


BBall %>% filter(position != "") %>% select(position,TotalInches) %>% ggplot(aes(x = TotalInches, color = position)) + geom_histogram() + facet_wrap(~position) + theme(legend.position = "None") + xlab("Inches")

BBall %>% filter(position != "") %>% ggplot(aes(x = TotalInches, color = position)) + geom_histogram() + facet_grid(~position) + theme(legend.position = "None") + xlab("Inches") + ylab("Count")
```


## Breakout: Create a column plot the compares the average height of all players that have come from Duke University, SMU and Texas Tech University.  
```{r}

BBall %>% filter(college == "Duke University" | college == "Southern Methodist University" | college == "Texas Tech University") %>% group_by(college) %>% dplyr::summarize(meanHeight = mean(TotalInches)) 

BBall %>% filter(college == "Duke University" | college == "Southern Methodist University" | college == "Texas Tech University") %>% group_by(college) %>% dplyr::summarize(meanHeight = mean(TotalInches)) %>% ggplot(aes(x = college, y = meanHeight, fill = college)) + geom_col() + theme(legend.position = "none") + ylab("Mean Height") + xlab("College")

BBall %>% filter(college == "Duke University" | college == "Southern Methodist University" | college == "Texas Tech University") %>% group_by(college) %>% dplyr::summarize(meanHeight = mean(TotalInches)) %>% ggplot(aes(x = college, y = meanHeight, fill = college)) + geom_col() + theme(legend.position = "none") + coord_cartesian(ylim = c(78.3, 78.6)) + ylab("Mean Height") + xlab("College")

# Conduct T-Test of mean height at Duke versus that of SMU

BBall_Duke = BBall %>% filter(college == "Duke University")

BBall_SMU = BBall %>% filter(college == "Southern Methodist University")

BBall_TTU = BBall %>% filter(college == "Texas Tech University")



t.test(BBall_TTU$TotalInches,BBall_SMU$TotalInches)

# There is not sufficient evidence to suggest that the mean height of basketball players from Duke is different from that of SMU (pvalue = .9203)

```

## Part 2 Soccer (Football)

```{r}
FIFA = read.csv(file.choose(),header = TRUE)
head(FIFA)

FIFA = FIFA %>% separate(Height, c("feet","inches"),sep = "'", remove = FALSE) %>% mutate(TotalInches = 12*as.numeric(feet) + as.numeric(inches)) %>% mutate(WeightNum = as.numeric(str_sub(FIFA$Weight,1,3)))
head(FIFA)

#Another way for weight:
FIFA = FIFA %>% separate(Height, c("feet","inches"),sep = "'", remove = FALSE) %>% mutate(TotalInches = 12*as.numeric(feet) + as.numeric(inches)) %>% mutate(WeightNum2 = as.numeric(str_replace(FIFA$Weight, "lbs", "")))
head(FIFA)

#or 

FIFA = FIFA %>% separate(Height, c("feet","inches"), sep = "'") %>% mutate(TotalInches = 12*as.numeric(feet) + as.numeric(inches)) %>% mutate(WeightNum3 = as.numeric(str_replace(FIFA$Weight, "...$", "")))
head(FIFA)



```

There appears to be a strong positive and linear relationship between height and weigth (as you would expect).  But is it the same for positions LM and LB?
```{r}
FIFA %>% ggplot(aes(x = WeightNum3, y = TotalInches)) + geom_point() + ggtitle("Height v. Weight")
FIFA %>% ggplot(aes(x = WeightNum3, y = TotalInches)) + geom_point() + ggtitle("Height v. Weight") + xlab("Weight (lbs)") + ylab("Height (inches)")

```

Visually, there is not a lot of evidence to suggest the  relationship is different. 
```{r}
FIFA %>% filter(Position == "LB" | Position == "LM") %>% ggplot(aes(x = WeightNum3, y = TotalInches, color = Position)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Height v. Weight (LM vs. LB)") +facet_wrap(~Position) + xlab("Weight (lbs)") + ylab("Height (inches)")

```

## Break Out FIFA
Assess a potential relationship between Wage vs Worth
```{r}
class(FIFA$Value)
class(FIFA$Wage)

#Isolate the number: Take Euro sign from the front and the K or M off the back
FIFA$Value2 = str_sub(FIFA$Value,2,-2)
FIFA$Wage2 = str_sub(FIFA$Wage,2,-2)

#Change to numeric
FIFA$Value2 = as.numeric(str_sub(FIFA$Value,2,-2))
FIFA$Wage2 = as.numeric(str_sub(FIFA$Wage,2,-2))

#Convert to same scale ... mult by 1000 if Value ends in a K and time 1000000 if it ends in a M
FIFA$KorM = str_sub(FIFA$Value,-1,-1)
FIFA$Value3 = if_else(FIFA$KorM == "K",FIFA$Value2 * 1000, FIFA$Value2*1000000,NaN)

#Plotting
FIFA %>% ggplot(aes(x = Value3, y = Wage2, col = Position)) + geom_point()
FIFA %>% ggplot(aes(x = Value3, y = Wage2, col = Position)) + geom_point() + facet_wrap(~Position)

```


## Part 3: Baby Names

```{r}
# 1A
# setwd()
df <- read.table(file.choose(), sep=";", header=FALSE, col.names = c("Name", "Gender", "BornThisYear"))
head(df)

#df$Gender = factor(df$Gender)

# 1B
summary(df)
str(df)

#1C
df$Name[grep("y{3}$", df$Name)] # There are others, of course; this regular expression does not even say it's at the end of the word

#1D 

y2016 <- df[-grep("y{3}$", df$Name),] # You could also hard-code which row it is, but I recommend this.
dim(y2016)

# 2A
y2015 <- read.table(file.choose(), sep=",", header=FALSE, col.names = c("Name", "Gender", "BornThisYear"))
head(y2015, n = 500)
# Notice that my columns are the same name.  If students have different names, the code below changes a little.


#2B
tail(y2015,n=2000)
# For 'interesting,' it could be a lot of things.  For one, all 10 are male.  All 10 start with the letter Z.  All are unpopular names. The data set is ordered by gender and then by BornThisYear and then alphabetically by Name.  

#Let's order the list only by Born this year agnostic to gender.
y2015 %>% arrange(desc(BornThisYear))

#2C
final<-merge(y2016, y2015, by=c("Name", "Gender"), all=FALSE) # There are other combinations of this.  Inner Join
head(final, n = 100)

#3A
final$Total <- final$BornThisYear.x + final$BornThisYear.y # Creating column
head(final)
sum(final$Total) # Figuring out everyone in it

#3B
a<-data.frame(final[order(-final$Total), c("Name", "Gender","Total")]) 
# You could also do tail here if you did ascending order.
head(a, n=10)

b<- final %>% arrange(desc(Total))
head(b, n=10)


#3C
# Lots of strategies here.  You could include only girls, omit only boys, use group bys, etc
girl<-head(a[a$Gender!="M",], n=10)
head(girl)
girl

girl2 = head(b %>% filter(Gender == "F"), n = 10)
head(girl2)
girl2

# 3D
girl<-girl[,c("Name", "Total")]
girl2 = girl %>% select(Name,Total)
head(girl)
head(girl2)

write.csv(girl2, "itsagirl.csv", row.names = FALSE)


# 4 Visualization

#girl$Name = droplevels(girl$Name)

girlnameplot = girl %>% ggplot(aes(x = factor(Name, levels = Name[order(Total)]), y = Total, fill = Name)) + geom_col() + ggtitle("Top 10 Girl Names!") + theme(legend.position = "none") + xlab("Name")

boy<-head(a[a$Gender == "M",], n=10)
boy2 = head(b %>% filter(Gender == "M"))

#boy$Name = droplevels(boy$Name)

boynameplot = boy %>% ggplot(aes(x = factor(Name, levels = Name[order(Total)]), y = Total, fill = Name)) + geom_col() + ggtitle("Top 10 Boy Names!") + theme(legend.position = "none") + xlab("Name")

library(patchwork)

girlnameplot + boynameplot


```

## Breakout Baby Names
Make a barplot of the frequency of female names that start with each letter of the alphabet.  In other words, for each letter of the alphabet, count the number of children that start with that letter and make a plot that conveys that information.
```{r}
Final_Fem = final %>% filter(Gender == "F")
Final_Fem$FirstLetter = str_sub(Final_Fem$Name,1,1)
Final_Fem %>% group_by(FirstLetter) %>% summarise(Total = sum(Total))
ByFirstLetter = Final_Fem %>% group_by(FirstLetter) %>% summarise(Total = sum(Total))
ByFirstLetter %>% ggplot(aes(x = FirstLetter, y = Total, fill = FirstLetter)) + geom_bar(stat = "Identity") + ggtitle("Frequency of Child Names by First Letter") + xlab("First Letter") + ylab("Frequency") + theme(legend.position = "none")
```

## Email / RegEx Exercise BreakOut
```{r}
grep("^[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}$", "bsadler@smu.edu")
```

## Links for updated names data:
https://www.ssa.gov/oact/babynames/
https://catalog.data.gov/dataset/popular-baby-names/resource/02e8f55e-2157-4cb2-961a-2aabb75cbc8b
https://github.com/aruljohn/popular-baby-names


## Cleaning Phone Numbers
```{r}
library(stringr)

x <- c(" (214) 555-0123 ", "214.555.0123", "+1 214 555 0123", "N/A")

clean_phone <- x %>%
  str_trim() %>%                                 # remove leading/trailing spaces
  str_replace_all("[^0-9]", "") %>%              # keep only digits
  str_replace("^1(?=\\d{10}$)", "") %>%          # drop leading country code 1 if present
  str_replace("^(\\d{3})(\\d{3})(\\d{4})$", "(\\1) \\2-\\3") # format

clean_phone

```


## Validating Emails
```{r}
library(stringr)

emails <- c("sydney.holder@smu.edu", "bob@", "a@b", "alice+test@gmail.com", "bad@@gmail.com")

# Simple “shape” check (not perfect, but good for QA screening)
is_valid <- str_detect(emails, "^[A-Za-z0-9._%+-]+@[A-Za-z0-9.-]+\\.[A-Za-z]{2,}$")

data.frame(email = emails, valid = is_valid)
```

