---
title: "Case Study 1 - Attrition"
author: "Isabella Westlauer"
date: "2026-03-08"
output: html_document
---
## Executive Summary
Employee attrition is a major challenge for organizations, as the loss of experienced employees can lead to increased hiring costs, decreased productivity, and disruptions in team performance. The goal of this study was to identify the factors most strongly associated with employee attrition and develop a predictive model that can help the organization identify employees who may be at higher risk of leaving. Using the provided employee dataset, I analyzed demographic, job satisfaction, compensation, and work-life balance variables, and engineered several additional features intended to capture employee stress, compensation dissatisfaction, and career progression.

Several factors emerged as strongly related to employee attrition. Employees with higher levels of burnout, characterized by working overtime combined with lower job satisfaction and work-life balance, were significantly more likely to leave the organization. Indicators of pay frustration, where employees had relatively low income compared to their level of involvement and responsibility, were also associated with higher attrition risk. Additionally, measures related to career stagnation, long commutes, and patterns of frequent job changes in the past were linked to an increased likelihood of attrition. These factors suggest that both workplace stress and perceived lack of career or compensation growth play important roles in employee turnover.

To evaluate attrition risk, I developed classification models using these engineered predictors. The final model achieved a balanced level of predictive performance when evaluated on a holdout test set. By adjusting the classification threshold, the model was able to reach approximately 60% sensitivity and 60% specificity, meaning it was reasonably effective at identifying employees who might leave while also limiting false alarms. This balance allows the model to serve as a practical early warning tool for HR teams seeking to intervene before attrition occurs.

Beyond the predictive model itself, the analysis highlights several broader insights. Attrition risk appears to be influenced by a combination of workload stress, compensation perceptions, and career development opportunities rather than any single factor alone. Employees experiencing multiple risk factors simultaneously—such as high workload, low satisfaction, and limited income growth—were particularly likely to leave. These findings suggest that organizations may reduce attrition by improving work-life balance, ensuring compensation aligns with employee contributions, and providing clearer pathways for career advancement.

Overall, this analysis demonstrates that employee attrition can be partially predicted using workplace and demographic indicators. By applying models such as the one developed in this study, organizations can proactively identify employees who may be at risk and take steps to improve retention through targeted engagement and support strategies.

## Link to YouTube presentation
https://www.youtube.com/watch?v=-fQ03kZ9Imw

## R Markdown
```{r}
### libraries for EDA and Initial Modeling
library(skimr)
library(gt)
library(tidyverse)
library(DataExplorer)
library(ggplot2)
library(plotly)
library(GGally)
```
## Data intake
```{r}
raw_data <- read.csv("CaseStudy1-data.csv")
```

## Prior Probability of Attrition
```{r}
# Count and Percent of Attrition from original dataset
overall_rate <- mean(raw_data$Attrition == "Yes") * 100 #16.09195%
attrition_positive = filter(raw_data, Attrition == "No")
dim(attrition_positive)[1] #730, 83.9%
attrition_negative = filter(raw_data, Attrition == "Yes")
dim(attrition_negative)[1] #140, 16.1%
```

## convert categorical data to numeric
```{r}
raw_data$Attrition_num <- ifelse(raw_data$Attrition == "Yes", 1, 0)
```

## Explore the spread of all features
```{r}
raw_data %>%
  plot_boxplot(by = "Attrition")
```

## Look at attrition by job role, department, marital status
```{r}
raw_data %>%
  group_by(Department) %>%
  #group_by(JobRole) %>%
  #group_by(MaritalStatus) %>%
  summarize(n = n(),
            attrition_rate = mean(Attrition == "Yes") * 100
  ) %>%
  arrange(desc(attrition_rate)) %>%
  ggplot(aes(x = reorder(Department, 
  # ggplot(aes(x = reorder(JobRole, 
  # ggplot(aes(x = reorder(MaritalStatus, 
                         attrition_rate), 
           y = attrition_rate,
           fill = Department)) +
          # fill = JobRole)) +
          # fill = MaritalStatus)) +
  geom_col() +coord_flip() +
  geom_text(aes(label = 
                  paste0(round(attrition_rate, 1), "%")),
            vjust = 0,
            hjust = 1.1,
            size = 5,
            fontface = "bold") +
  labs(
    title = "Attrition Rate by Department",
    # title = "Attrition Rate by Job Role",
    # title = "Attrition Rate by Marital Status",
    x = NULL,
    y = "Attrition Rate (%)"
  ) +
  theme_minimal(base_size = 14)+
  theme(legend.position = "none")
```

```{r}
raw_data$OverTime_num <- ifelse(raw_data$OverTime == "Yes", 1, 0)
```

## Plot attrition percentage for job involvement, stock option level
```{r}
raw_data %>%
  # group_by(JobInvolvement) %>%
  group_by(StockOptionLevel) %>%
  summarise(
    n = n(),
    attrition_rate = mean(Attrition == "Yes") * 100
  ) %>%
  # arrange(JobInvolvement) %>%
  arrange(StockOptionLevel) %>%
  # ggplot(aes(x = factor(JobInvolvement),
  ggplot(aes(x = factor(StockOptionLevel),
           y = attrition_rate)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = 
                  paste0(round(attrition_rate, 1), "%")),
            vjust = 1,
            size = 5,
            fontface = "bold") +
  geom_hline(yintercept = overall_rate,
             linetype = "dashed",
             color = "red",
             size = 1) +
  labs(
    # title = "Attrition Rate by Job Involvement",
    title = "Attrition Rate by Stock Option Level",
    # x = "Job Involvement",
    x = "Stock Option Level",
    y = "Attrition Rate (%)"
  ) +
  theme_minimal(base_size = 14)
```
## Try the ggbetweenstats to visualize total working years
```{r}
# Violin plot, box plot, scatter plot, and density plot
library(ggstatsplot)
plt <- ggbetweenstats(
  data = raw_data,
  x = Attrition,
  y = TotalWorkingYears
)
plt
```

## Visualize attrition rate by overtime
```{r}
raw_data %>%
  group_by(OverTime) %>%
  summarise(
    n = n(),
    attrition_rate = mean(Attrition == "Yes") * 100
  ) %>%
  ggplot(aes(x = OverTime,
           y = attrition_rate,
           fill = OverTime)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(attrition_rate, 1), "%")),
            vjust = 1,
            size = 5,
            fontface = "bold") +
  geom_hline(yintercept = overall_rate,
             linetype = "dashed",
             color = "red",
             size = 1) +
  labs(
    title = "Attrition Rate by Overtime Status",
    subtitle = paste("Overall Attrition Rate:", 
                     round(overall_rate, 1), "%"),
    x = "Overtime",
    y = "Attrition Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("Yes" = "#e88631",
                               "No"  = "#7c3896")) + 
  theme(legend.position = "none")
```

## Correlation Heatmap, Bar Plot, GG Pairs of Selected Features
```{r}
raw_data %>% select(
  "Attrition",
  "JobInvolvement",
  # "JobLevel",
  # "StockOptionLevel",
  "MonthlyIncome",
  # "TotalWorkingYears",
  # "MaritalStatus",
  # "JobRole",
  "OverTime"
# ) %>% plot_correlation()
# ) %>% plot_bar()
) %>% ggpairs(aes(color = Attrition))
```

## Classification Models
```{r}
library(class)
library(caret)
library(e1071) # Naive Bayes
```

### Change response variable to a factor and make "Yes" the positive class
```{r}
# Set attrition as factor and check current levels:
raw_data$Attrition <- as.factor(raw_data$Attrition)
levels(raw_data$Attrition) #no, yes
raw_data$Attrition <- relevel(raw_data$Attrition, "Yes")
levels(raw_data$Attrition) #yes, no
```

## Feature Engineering
This should have been done after train/test split
```{r}
# Split monthly income into four levels based on quartile
income_breaks <- quantile(raw_data$MonthlyIncome,
                          probs = c(0, 0.25, 0.5, 0.75, 1))
raw_data$income_quartile_labels <- cut(raw_data$MonthlyIncome,
                               breaks = income_breaks,
                               include.lowest = TRUE,
                               labels = c("Q1_Low",
                                          "Q2_MidLow",
                                          "Q3_MidHigh",
                                          "Q4_High"))
raw_data$income_quartile_num <- as.numeric(raw_data$income_quartile_labels)

raw_data$MaritalStatusSingle <- ifelse(raw_data$MaritalStatus == "Single", 1, 0)

# Burnout Index: Overtime + (4- Work Life Balance) + (4- Job Satisfaction)
raw_data$burnout_index <-
  raw_data$OverTime_num + 
  (4-raw_data$WorkLifeBalance) + 
  (4-raw_data$JobSatisfaction)

# Pay Frustration: (Job Level + Job Involvement) / Monthly Income
raw_data$pay_frustration <-
  (raw_data$JobLevel + raw_data$JobInvolvement)*2000 /
  raw_data$MonthlyIncome

# Career Stagnation: Years Since Last Promotion / (Years at Company +1)
raw_data$career_stagnation <-
  raw_data$YearsSinceLastPromotion /
  (raw_data$YearsAtCompany + 1)

# Commute Pain: Distance from Home * Overtime
raw_data$commute_pain <-
  raw_data$DistanceFromHome *
  raw_data$OverTime_num

raw_data$job_hopping_rate <-
  raw_data$MaritalStatusSingle *
  (raw_data$NumCompaniesWorked / 
  raw_data$Age)

raw_data$compensation_unfairness <- 
  raw_data$MonthlyIncome * 
  raw_data$JobInvolvement *
  (1+raw_data$OverTime_num)
```

## Train/Test Split
```{r}
#split and seed can be changed
set.seed(60)
split = .7

obs = dim(raw_data)[1]
trainIndices = sample(seq(1:obs),round(split*obs))
trainSet = raw_data[trainIndices,]
testSet = raw_data[-trainIndices,]
```

## KNN model (not optimized)
```{r}
# tunable threshold, hyperparameter
classification_threshold = 0.2
k_value = 19
```

# Feature selection
```{r}
train_knn <- scale(trainSet[,c(
                              "burnout_index"
                               # ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              # ,"commute_pain"
                              # ,"job_hopping_rate"
                              # ,"compensation_unfairness"
                              )])

test_knn  <- scale(testSet[,c(
                              "burnout_index"
                              # ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              # ,"commute_pain"
                              # ,"job_hopping_rate"
                              # ,"compensation_unfairness"
                              )],
                   center = attr(train_knn,"scaled:center"),
                   scale  = attr(train_knn,"scaled:scale"))
```

```{r}
preds <- knn(train_knn,
             test_knn,trainSet$Attrition,
             k = k_value, 
             prob = TRUE)

prob <- attr(preds, "prob")

prob_yes <- ifelse(preds == "Yes", prob, 1 - prob)

preds2 <- ifelse(prob_yes > classification_threshold, "Yes", "No")
preds2 <- factor(preds2, levels = levels(testSet$Attrition))

cm <- preds2 %>%
  table(testSet$Attrition) %>%
  confusionMatrix()

cm$overall[1] #accuracy
cm$byClass[1] #sensitivity
cm$byClass[2] #specificity
```
## Tune the hyperparameter k
We find that k=19 has the highest sensitivity. 

```{r}
# I only want to pick an odd numbered k
set.seed(60)
numks = 120 # For each of the 60 values of k that we are trying
iterations = 30 # obtain an accuracy statistic for 30 different train test splits
split = .7
obs = dim(raw_data)[1]

# I only want the highest sensitivity
masterSens = matrix(nrow = iterations, ncol = numks)

for(j in 1:iterations)
{

  trainIndices = sample(seq(1:obs),round(split*obs))
  trainSet = raw_data[trainIndices,]
  testSet = raw_data[-trainIndices,]
  
  train_knn <- scale(trainSet[,c(
                              "burnout_index"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              )])

  test_knn  <- scale(testSet[,c(
                              "burnout_index"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              )],
                   center = attr(train_knn,"scaled:center"),
                   scale  = attr(train_knn,"scaled:scale"))
  for(i in seq(1, numks, by = 2))
    {
      preds <- knn(train_knn,
               test_knn,trainSet$Attrition,
               k = i, 
               prob = TRUE)
  
      prob <- attr(preds, "prob")
      prob_yes <- ifelse(preds == "Yes", prob, 1 - prob)
      preds2 <- ifelse(prob_yes > classification_threshold, "Yes", "No")
      preds2 <- factor(preds2, levels = levels(testSet$Attrition))
      
      cm <- preds2 %>%
        table(testSet$Attrition) %>%
        confusionMatrix()
      masterSens[j,i] = cm$byClass[1]
    }
}

MeanSens = colMeans(masterSens)
plot(seq(1,numks,1),MeanSens, type = "l", ylim = c(.5,1))
which.max(MeanSens) # which k value has the highest sensitivity? k=13
max(MeanSens)
```
## Tune the classification threshold
```{r}
train_knn <- scale(trainSet[,c(
                              "burnout_index"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              )])

test_knn  <- scale(testSet[,c(
                              "burnout_index"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              )],
                   center = attr(train_knn,"scaled:center"),
                   scale  = attr(train_knn,"scaled:scale"))
preds <- knn(train_knn,
             test_knn,trainSet$Attrition,
             k = 19, 
             prob = TRUE)
prob <- attr(preds, "prob")
prob_yes <- ifelse(preds == "Yes", prob, 1 - prob)
```
```{r}
# zero in on the threshold range:
# thresholds <- seq(0.45, 0.05, by = -0.05)
# thresholds <- seq(0.3, 0.1, by = -0.01)
thresholds <- seq(0.11, 0.09, by = -0.001)
length(thresholds)

accs = data.frame(
  accuracy = numeric(length(thresholds)),
  sensitivity = numeric(length(thresholds)),
  specificity = numeric(length(thresholds)),
  yes_threshold = thresholds
)

for (j in seq_along(thresholds)) {

  t <- thresholds[j]

  preds2 <- ifelse(prob_yes > t, "Yes", "No")
  preds2 <- factor(preds2, levels = levels(testSet$Attrition))

  cm <- preds2 %>%
    table(testSet$Attrition) %>%
    confusionMatrix()

  accs$accuracy[j] <- cm$overall[1]
  accs$sensitivity[j] <- cm$byClass[1]
  accs$specificity[j] <- cm$byClass[2]
}

max_sens <- max(accs$sensitivity)
best_sens <- accs$sensitivity[accs$sensitivity == max_sens]
best_acc <-accs$accuracy[accs$sensitivity == max_sens]
best_spec<-accs$specificity[accs$sensitivity == max_sens] 

#get the highest sensitivity where specificity is at least 0.6
valid <- accs %>%
  filter(specificity >= 0.6)
best_row <- valid %>%
  slice_max(sensitivity, n = 1)
best_threshold <- max(best_row$yes_threshold)
best_sens <- max(best_row$sensitivity)
best_spec <- max(best_row$specificity)
best_acc <- max(best_row$accuracy)

best_threshold
best_sens
best_spec
best_acc
```

## Plot the threshold accuracies
```{r}
library(tidyr)
library(ggplot2)

accs_long <- accs |>
  pivot_longer(
    cols = c(sensitivity, specificity),
    names_to = "metric",
    values_to = "value"
  )

label_text <- paste0(
  "Threshold = ", round(best_threshold, 2),
  "\nSensitivity = ", round(best_sens, 2),
  "\nSpecificity = ", round(best_spec, 2)
)

ggplot(accs_long,
       aes(x = yes_threshold,
           y = value,
           color = metric)) +

  geom_line(size = 1.3) +

  geom_point(size = 3) +

  geom_hline(yintercept = 0.6,
             linetype = "dashed",
             color = "gray40") +
  geom_vline(xintercept = best_threshold,
           linetype = "dashed",
           color = "darkgreen",
           size = 1) +
  
    annotate("text",
           x = best_threshold + 0.02,
           y = 0.8,
           label = label_text,
           hjust = 0,
           size = 4,
           color = "black") +

  scale_color_manual(values = c("steelblue","firebrick")) +

  labs(
    title = "Effect of Classification Threshold on Model Performance",
    x = "Probability Threshold for Predicting Attrition = Yes",
    y = "Model Performance",
    color = "Metric"
  ) +

  theme_minimal(base_size = 14)

```
## Test the optimized KNN model
```{r}
set.seed(60)
split = .7

obs = dim(raw_data)[1]
trainIndices = sample(seq(1:obs),round(split*obs))
trainSet = raw_data[trainIndices,]
testSet = raw_data[-trainIndices,]
classification_threshold = 0.104
k_value = 19
train_knn <- scale(trainSet[,c(
                              "burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                              )])

test_knn  <- scale(testSet[,c(
                              "burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                              )],
                   center = attr(train_knn,"scaled:center"),
                   scale  = attr(train_knn,"scaled:scale"))
preds <- knn(train_knn,
             test_knn,trainSet$Attrition,
             k = k_value, 
             prob = TRUE)

prob <- attr(preds, "prob")

prob_yes <- ifelse(preds == "Yes", prob, 1 - prob)

preds2 <- ifelse(prob_yes > classification_threshold, "Yes", "No")
preds2 <- factor(preds2, levels = levels(testSet$Attrition))

cm <- preds2 %>%
  table(testSet$Attrition) %>%
  confusionMatrix()
cm


```

## Naive Bayes Model
```{r}
model <- naiveBayes(trainSet[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                              # ,"compensation_unfairness"
                                 )],
                    trainSet$Attrition)

pred_probs <- predict(model,
                 testSet[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                              # ,"compensation_unfairness"
                             )],
                 type = "raw")
# 
# 
# pred_class <- ifelse(pred_probs[, "Yes"] > 0.1, "Yes", "No")
# CM_NB <- confusionMatrix(factor(pred_class, levels=c("No","Yes")),
#                 testSet$Attrition)
# CM_NB$overall[1]
# CM_NB$byClass[1]
# CM_NB$byClass[2]
```

## Optimizing the NB model
```{r}
prob_yes <- pred_probs[, "Yes"]

# zero in on the threshold range
# thresholds <- seq(0.5, 0.01, by = -0.01)
# thresholds <- seq(0.3, 0.05, by = -0.01)
thresholds <- seq(0.1, 0.01, by = -0.001)
# length(thresholds)

accs_nb <- data.frame(
  accuracy = numeric(length(thresholds)),
  sensitivity = numeric(length(thresholds)),
  specificity = numeric(length(thresholds)),
  yes_threshold = thresholds
)

for (j in seq_along(thresholds)) {

  t <- thresholds[j]

  preds <- ifelse(prob_yes > t, "Yes", "No")
  preds <- factor(preds, levels = levels(testSet$Attrition))

  cm <- confusionMatrix(preds, testSet$Attrition)

  accs_nb$accuracy[j] <- cm$overall[1]
  accs_nb$sensitivity[j] <- cm$byClass[1]
  accs_nb$specificity[j] <- cm$byClass[2]
}

valid_nb <- accs_nb %>%
  filter(specificity >= 0.6)

best_nb <- valid_nb %>%
  slice_max(sensitivity, n = 1)

best_nb
```

## Visualize the comparison between models
```{r}
accs$model <- "KNN"
accs_nb$model <- "Naive Bayes"
all_accs <- rbind(accs, accs_nb)

all_long <- all_accs |>
  pivot_longer(
    cols = c(sensitivity, specificity),
    names_to = "metric",
    values_to = "value"
  )
ggplot(all_long,
       aes(x = yes_threshold,
           y = value,
           color = metric)) +

  geom_line(size = 1.2) +
  geom_point(size = 2) +
  
  geom_hline(yintercept = 0.6,
             linetype = "dashed",
             color = "gray40") +

  facet_wrap(~model) +

  scale_color_manual(values = c("steelblue","firebrick")) +

  labs(
    title = "Threshold Tuning Comparison: k-NN vs Naive Bayes",
    x = "Probability Threshold for Predicting Attrition = Yes",
    y = "Model Performance",
    color = "Metric"
  ) +

  theme_minimal(base_size = 14)
```

## Test optimized NB model
```{r}
model <- naiveBayes(trainSet[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                                 )],
                    trainSet$Attrition)

pred_probs <- predict(model,
                 testSet[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                             )],
                 type = "raw")


pred_class <- ifelse(pred_probs[, "Yes"] > 0.066, "Yes", "No")
CM_NB <- confusionMatrix(factor(pred_class, levels=c("No","Yes")),
                testSet$Attrition)
CM_NB
```

## Predict on unlabeled data
```{r}
study_set <- read.csv("CaseStudy1-data.csv")
# study_set$Attrition_num <- ifelse(study_set$Attrition == "Yes", 1, 0)

study_set$Attrition <- factor(study_set$Attrition,
                              levels = c("No","Yes"))

levels(study_set$Attrition)
study_set$Attrition <- relevel(study_set$Attrition, "Yes")
levels(study_set$Attrition)
# Split monthly income into four levels based on quartile
income_breaks <- quantile(study_set$MonthlyIncome,
                          probs = c(0, 0.25, 0.5, 0.75, 1))
study_set$income_quartile_labels <- cut(study_set$MonthlyIncome,
                               breaks = income_breaks,
                               include.lowest = TRUE,
                               labels = c("Q1_Low",
                                          "Q2_MidLow",
                                          "Q3_MidHigh",
                                          "Q4_High"))
study_set$income_quartile_num <- as.numeric(study_set$income_quartile_labels)

study_set$MaritalStatusSingle <- ifelse(study_set$MaritalStatus == "Single", 1, 0)

# Burnout Index: Overtime + (4- Work Life Balance) + (4- Job Satisfaction)
study_set$OverTime_num <- ifelse(study_set$OverTime == "Yes", 1, 0)
study_set$burnout_index <-
  study_set$OverTime_num + 
  (4-study_set$WorkLifeBalance) + 
  (4-study_set$JobSatisfaction)

# Pay Frustration: (Job Level + Job Involvement) / Monthly Income
study_set$pay_frustration <-
  (study_set$JobLevel + study_set$JobInvolvement)*2000 /
  study_set$MonthlyIncome

# Career Stagnation: Years Since Last Promotion / (Years at Company +1)
study_set$career_stagnation <-
  study_set$YearsSinceLastPromotion /
  (study_set$YearsAtCompany + 1)

# Commute Pain: Distance from Home * Overtime
study_set$commute_pain <-
  study_set$DistanceFromHome *
  study_set$OverTime_num

study_set$job_hopping_rate <-
  study_set$MaritalStatusSingle *
  (study_set$NumCompaniesWorked / 
  study_set$Age)

study_set$compensation_unfairness <- 
  study_set$MonthlyIncome * 
  study_set$JobInvolvement *
  (1+study_set$OverTime_num)

competition_model <- naiveBayes(study_set[, c("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate"
                                 )],
                    study_set$Attrition)



competition_set = read.csv("CaseStudy1CompSet No Attrition.csv")



#burnout index
competition_set$OverTime_num <- ifelse(competition_set$OverTime == "Yes", 1, 0)
competition_set$burnout_index <-
  competition_set$OverTime_num + 
  (4-competition_set$WorkLifeBalance) + 
  (4-competition_set$JobSatisfaction)
#
# Pay Frustration: (Job Level + Job Involvement) / Monthly Income
competition_set$pay_frustration <-
  (competition_set$JobLevel + competition_set$JobInvolvement)*2000 /
  competition_set$MonthlyIncome

# Career Stagnation: Years Since Last Promotion / (Years at Company +1)
competition_set$career_stagnation <-
  competition_set$YearsSinceLastPromotion /
  (competition_set$YearsAtCompany + 1)

# Commute Pain: Distance from Home * Overtime
competition_set$commute_pain <-
  competition_set$DistanceFromHome *
  competition_set$OverTime_num

competition_set$MaritalStatusSingle <- ifelse(competition_set$MaritalStatus == "Single", 1, 0)
competition_set$job_hopping_rate <-
  competition_set$MaritalStatusSingle *
  (competition_set$NumCompaniesWorked / 
  competition_set$Age)

competition_set$compensation_unfairness <- 
  competition_set$MonthlyIncome * 
  competition_set$JobInvolvement *
  (1+competition_set$OverTime_num)

competition_set$income_quartile_labels <- cut(competition_set$MonthlyIncome,
                               breaks = income_breaks,
                               include.lowest = TRUE,
                               labels = c("Q1_Low",
                                          "Q2_MidLow",
                                          "Q3_MidHigh",
                                          "Q4_High"))
competition_set$income_quartile_num <- 
  as.numeric(competition_set$income_quartile_labels)
library(tidyverse)
competition_set_selected <- competition_set %>%select("burnout_index"
                              ,"pay_frustration"
                              ,"income_quartile_num"
                              ,"career_stagnation"
                              ,"commute_pain"
                              ,"job_hopping_rate")
competition_probs <- predict(competition_model,
                 competition_set_selected,
                 type = "raw")
competition_output <- data.frame(
  Attrition_Probability = competition_probs[,"Yes"]
)
write.csv(competition_output, "Case1PredictionsWestlauer Attrition.csv", row.names = FALSE)
```
