---
title: "unit 2 practice"
author: "Isabella Westlauer"
date: "2026-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggthemes)
```

##Get a summary of the data
```{r}
library(readr)
PlayersBBall <- read_csv("C:/Users/isabe/Desktop/MSDS6306/MSDS_6306_DDS/MSDS_6306_Doing-Data-Science/Unit 2/PlayersBBall.csv")
View(PlayersBBall)
summary(PlayersBBall)
```
##Distribution of players by position  
```{r}

ggplot(PlayersBBall, aes(x = position)) +
  ggtitle("Distribution of players by position")+ 
  geom_bar()

```  

##Distribution of weights of forwards and centers  
###Forwards  
```{r}
PlayersBBall[PlayersBBall$position == "F",] %>% 
  ggplot(aes(x= weight)) + 
  ggtitle("Distribution of Weights of Forwards") +
  xlab("Weight (lbs)") +
  ylab("count")+
  geom_histogram(fill = "blue", color = "black")

forwards <- PlayersBBall[PlayersBBall$position == "F",]
mean(forwards$weight, na.rm = TRUE)
```

###Centers  
```{r}
PlayersBBall[PlayersBBall$position == "C",] %>% 
  ggplot(aes(x= weight)) + 
  ggtitle("Distribution of Weights of Centers") +
  xlab("Weight (lbs)") +
  geom_histogram(fill = "green", color = "black")  

centers <- PlayersBBall[PlayersBBall$position == "C",]
mean(centers$weight, na.rm = TRUE)
```

##Distribution of player height, comparing forwards and centers  
```{r}
# uh-oh! height in the csv is not formatted as a float

PlayersBBall <- PlayersBBall %>%
  separate(height, into = c("feet", "inches"), sep = "-", convert = TRUE) %>%
  mutate(height_float = feet + inches / 12)

head(PlayersBBall[, c("feet", "inches", "height_float")]) # Quick inspection
```
###Height of forwards  
```{r}
PlayersBBall[PlayersBBall$position == "F",] %>%
  ggplot(aes(x = height_float)) +
  ggtitle("Distribution of Height (Feet) of Forwards")+
  xlab("Height (feet)") +
  geom_density(fill = "blue")

forwards <- PlayersBBall[PlayersBBall$position == "F",]
mean(forwards$height_float, na.rm = TRUE)
```
###Height of centers  
```{r}
PlayersBBall[PlayersBBall$position == "C",] %>%
  ggplot(aes(x = height_float)) +
  ggtitle("Distribution of Height (Feet) of Centers")+
  xlab("Height (feet)") +
  geom_density(fill = "green")

centers <- PlayersBBall[PlayersBBall$position == "C",]
mean(centers$height_float, na.rm = TRUE)
```

##Height comparison of Forwards and Centers
```{r}

ggplot() +
  geom_density(
    data = PlayersBBall[grepl("F", PlayersBBall$position), ],
    aes(x = height_float),
    fill = "blue",
    alpha = 0.4
  ) +
  geom_density(
    data = PlayersBBall[grepl("C", PlayersBBall$position), ],
    aes(x = height_float),
    fill = "green",
    alpha = 0.4
  ) +
  labs(
    title = "Height Distribution of Forwards vs Centers",
    x = "Height (feet)",
    y = "Density"
  ) +
  theme_minimal()

```


##Density plot of the heights of players in all positions  
```{r}
density_plot_height = ggplot(PlayersBBall, aes(x = height_float, color = position)) +
  ggtitle("Distribution of Height (Feet) of Players by Position") +
  xlab("Height (feet)")+
  geom_density(linewidth = 0.75)

ggplotly(density_plot_height)
```

##Facet grid for height and weight by position
```{r}
#create a smaller dataframe with just height and weight and position
player_size <- PlayersBBall %>%
  select(position, height_float, weight)

# Make the data longer so we can use height and weight as measures in a facet grid
player_size_long <- player_size %>%
  pivot_longer(
    cols = c(height_float, weight),
    names_to = "measure",
    values_to = "value"
  )

# head(player_size_long)

# This is bad because the x scale doesn't work for both height and weight
# ggplot(data = player_size_long, aes(fill = position)) + 
# geom_histogram(mapping = aes(x = value)) +
# facet_grid(measure~position)

# this doesn't look good because the x labels are wrong
# ggplot(player_size_long, aes(x = value, fill = position)) + 
#   geom_histogram() +
#   facet_grid(measure ~ position, scales = "free_x") +
#   scale_x_continuous(breaks = seq(0,500, by = 100)) +
#   theme_minimal()

#Switch the measure and position!
ggplot(player_size_long, aes(x = value, fill = position)) + 
  geom_histogram() +
  facet_grid(position ~ measure, scales = "free_x") +
  theme_minimal()

```

##Height vs weight, does the correlation change based on position?  

The height and weight of the players have a positive correlation. 
The relationship doesn't change much between different positions.
```{r}
ggplot(PlayersBBall,aes(x = height_float, y = weight, color = position)) + 
  ggtitle("Height v Weight of Players by Position") + 
  xlab("Height (feet)")+
  ylab("Weight (lbs)")+
  geom_point()
```

Use linear regression to see if the slopes don't match up.
The slopes all look nearly the same.
```{r}
ggplot(PlayersBBall,aes(x = height_float, y = weight, color = position)) +
  ggtitle("Height v Weight of Players by Position (using linear regression)") +
  xlab("Height (feet)")+
  ylab("Weight (lbs)")+
  geom_point() +
  geom_smooth(method = "lm")
```

Use facet wrap to get a better look
```{r}
ggplot(data = PlayersBBall) + 
ggtitle("Height v Weight of Players by Position (lm and facet wrap)") +
xlab("Height (feet)")+
ylab("Weight (lbs)")+
geom_point(mapping = aes(x = height_float, y = weight, color = position)) +
geom_smooth(mapping = aes(x = height_float, y = weight, linetype = position)) +
facet_wrap(~position)

```


##Distribution of player height by start year  

- Assuming I can use year_start for the year, because a player's height wouldn't change.
- Using boxplots because I don't want outliers to affect the mean of the height  
```{r}
# This doesn't work because the x labels are all overlapping and look bad
# ggplot(PlayersBBall, aes(x = factor(year_start), y = height_float)) +
#   geom_boxplot() +
#   labs(
#     title = "Distribution of Player Heights by Start Year",
#     x = "Start Year",
#     y = "Height (feet)"
#   ) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Add breaks to the x scale so labels show every 10 years in the range  
```{r}
# Here's the fix
p <- ggplot(PlayersBBall, aes(x = year_start, y = height_float)) +
  geom_boxplot(aes(group = year_start)) +
  scale_x_continuous(breaks = seq(min(PlayersBBall$year_start),
                                  max(PlayersBBall$year_start),
                                  by = 10)) +
  labs(
    title = "Distribution of Player Heights by Start Year",
    x = "Start Year",
    y = "Height (feet)"
  ) +
  # theme_minimal()
  # theme_stata()
  theme_economist()
  # theme_clean()
  # theme_fivethirtyeight()
  # theme_tufte()
  # theme_solarized()
  # theme_pander()
  # theme_excel_new()
  # theme_wsj()

ggplotly(p)
```

##Use plotly library to show height vs weight vs start year with color indicating position  
```{r}
library(plotly)
p <- plot_ly(PlayersBBall, 
             x = ~height_float, y = ~weight, z = ~year_start, 
             color = ~position) %>%
   add_markers() %>%
   layout(scene = list(title = "Player Height vs Weight vs Start Year",
                       xaxis = list(title = 'Height (feet)'),
                      yaxis = list(title = 'Weight (lbs)'),
                      zaxis = list(title = 'Start Year')))
ggplotly(p)
```
